Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-13 papers/wdmerger_I/paper.tex
Updates in response to Mike's first pass

The latest developments have resulted in some possible areas of convergence.
\cite{pakmor:2012} performed a merger scenario
with a $1.1\ \msolar$ and $0.9\ \msolar$ setup, with better treatment
of the initial conditions, and indeed found that the merger process
happened over more than ten orbits. Nevertheless, they still determined
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} was also able to find a detonation
in a similarly massive system. Very recently, \cite{kashyap:2015} 
found that a detonation occurred in a similarly massive system. Notably,
the detonation occurred self-consistently and did not need to be inserted 
by hand. \cite{dan:2012} and \cite{dan:2014} performed a large sweep 
of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many
would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. For now there is preliminary support for the hypothesis
that systems with helium shells (or helium WDs), and very massive carbon-oxygen binaries,
could robustly lead to events resembling SNe Ia.

@@

This is the first in a series of papers designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
%\hyperref[sec:Numerical Methodology]{Section \ref*{sec:Numerical Methodology}}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:initial_models} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to demonstrate that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

The Euler equations for hydrodynamics (in the absence of source terms) in conservative form are:
\begin{align}
  \frac{\partial \rho}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}) \label{eq:euler_density}\\
  \frac{\partial \rho \mathbf{u}}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}\mathbf{u}) - \bm{\nabla}p \label{eq:euler_momentum}\\
  \frac{\partial \rho E}{\partial t} &= -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u}). \label{eq:euler_energy}
\end{align}
Here $\rho$ is the mass density, $\mathbf{u}$ is the fluid velocity
vector, $p$ is the pressure, and $E = \mathbf{u}^2 / 2 + e$ is the
total specific energy, where $e$ is the internal (thermal) specific
energy (energy per unit mass).

@@

A related issue that required a code improvement is that in cases of
large density gradients such as the edge of a star, it is possible to
generate negative densities in zones even with the more strongly
limited PPM. This can occur if a region of large density is moving
away from an ambient zone at relatively large speeds; then the net
density flux in the ambient zones can be large enough to unphysically
drag the density below zero. In practice, this will occur at the
trailing edge of a star that is moving across a grid. In such a
situation, there are two main approaches one could take: either
explicitly introduce a positivity-guaranteeing diffusive flux, or
reset the characteristics of the affected zone. We choose the latter
approach. Even though it is non-conservative, it preserves a
characteristic we value, which is to keep the edge of the stars
relatively sharp, as they physically should be. Since the mass of the
affected zones is typically already fairly low, this should not
seriously affect the energy conservation properties of our
simulation. Our strategy for a reset is as follows: when the density of 
a zone is below a pre-determined density floor (which is typically 
$10^{-5}\ \text{g cm}^{-3}$ for our stellar simulations), we look
at all adjacent zones and find the zone with the highest density.
If it is above the density floor, then we set the field values 
(density, momentum, energy, and temperature) of the
resetted zone to be exactly equal to the field values of this 
adjacent zone. If no adjacent zone reaches the density floor, then
the zone is set to the density floor, and given a temperature equal 
to the temperature floor for our simulations (which is typically 
$10^{5}\ \text{K}$ for our stellar simulations). We then recompute 
the thermodynamics to be consistent with these values. The 
velocity of the zone is set to zero. This latter approach will only
occur in very rare situations, and is mainly there as a safeguard.

\castro's approach to adaptive mesh refinement, based on its underlying
\boxlib\ framework, is to refine zones based on certain user-specified
criteria that tag regions of interest for higher spatial
resolution. Data is represented on one of a number of AMR levels,
where each level corresponds to a set of zones at the same resolution,
which covers a subset of the domain covered by the level immediately
below it. We typically call the level 0 grid the \textit{coarse} grid,
which has the lowest spatial resolution. Each finer, higher-level grid
has a higher resolution than the grid below it by some integer factor
$N$, which is restricted to be $N = 2\ \text{or}\ 4$ in the code. The
zones are strictly contained within the rectangular extent of the
underlying coarser zones (the code is not restricted to representing
only Cartesian geometries, but we use a Cartesian mesh with uniform
spacing in each dimension for the present study). For the time
evolution of the AMR system we use subcycling, where each AMR level is
advanced at a different timestep and a correction step is applied at
the end to synchronize the various levels. Generally the number of
subcycled timesteps is equal to the jump in refinement between levels,
so for example on a grid with three levels and two jumps of four in
refinement, the level 2 zones will have 16 times higher spatial
resolution than the coarse grid and there will be 16 level 2 timesteps
per level 0 timestep.

@@

There are two ways that these source terms enter the hydrodynamics update.
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. The details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
However, we believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

@@

In a typical simulation we place the boundaries of the domain far
enough away from the region containing most of the mass that some
method of approximation to this direct summation is justified. Many
approaches exist in the literature. The original release of \castro\
featured the crudest possible approximation: a monopole prescription,
where the boundary values were computed by summing up all the mass on
the domain and treating it as a point source at the domain
center. This is exactly correct only for a spherically symmetric mass
distribution, and therefore is best suited for problems like
single-star Type Ia supernova simulations that employ
self-gravity. However, for a problem like that of a binary star system
with significant departures from spherical symmetry, this assumption
fails to produce accurate boundary values. This results in a
significant drift of the center of the mass of the system over time,
which will be discussed in \autoref{sec:kepler}.

@@

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, loaded onto the grid by the method described 
in \autoref{sec:initial_models}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by \autoref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we will call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in \autoref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. In \autoref{sec:kepler} we 
see that there are no gains to be had by increasing the accuracy further. 
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

@@

We can solve one of these two sources of error by changing the density on  
the grid so that the total mass $M$ is actually what we intend it to be.
The resolution study for this case is also
plotted in \autoref{fig:gravity_convergence}. We still obtain
convergence slightly better than second-order, indicating that we 
have not eliminated the geometrical problem of the mass distribution changing.

@@

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers. We use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations. However, a slight 
difference is that the Coriolis force for each velocity component is coupled to other velocity 
components. If the rotation is about the $z$-axis, then the discrete update to 
$v_x^{n+1}$ depends on the value of $v_y^{n+1}$, and vice versa. If we fix the value of 
the time-level $n+1$ quantities after coming out of the hydrodynamics update, there 
would be a slight inconsistency between the $x$ and $y$ components of the velocity.

@@

In all simulations performed in a rotating reference frame, we transform all relevant
quantities back to the inertial reference frame when reporting them. In particular,
for every zone we adjust the momenta and kinetic energy to account for rotation at 
frequency $\omega$. When plotting the locations of 
stars on the grid, we rotate them by an angle corresponding to the amount of time that has 
passed since the simulation started.

%==========================================================================
% Initial Models and Problem Setup
%==========================================================================

\section{Initial Models and Problem Setup}
\label{sec:initial_models}

@@

We map the 1D model onto the 3D Cartesian grid by taking density,
temperature, and composition as the independent variables,
interpolating these to the cell centers, and then calling the equation
of state to initialize the remaining terms. It is possible to interpolate
instead by using pressure instead of temperature, as pressure is more 
closely related to hydrostatic balance, but the EOS we use is so 
insensitive to temperature that this mapping can result in large 
deviations from the isothermal assumption we started with.  The 
interpolation process divides each zone into $n_{\text{sub}}$ 
sub-zones of equal volume for
the purpose of sampling the 1D model, and the sub-zones are added
together to obtain the full zone's characteristics. This
sub-grid-scale interpolation is useful especially near the edge of the star,
where the density falls off rapidly with radius. Typically we take 
$n_{\text{sub}} = 4$.

@@

@@

@@

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the origin, 
and evolved it for 200 seconds. Our diagnostic of choice is the effective 
radius of the star, determined by the volume of the grid that has a density 
greater than $10^3\ \text{g cm}^{-3}$ (see \autoref{sec:initial_models} 
for details on this measure). This choice of density is intended to 
mark a reasonable outer edge to the star that is not immediately susceptible 
to the numerical errors prevalent near the physical edge of the star. 
\autoref{fig:single_star_static_radius} shows our results at various resolutions. 
As expected, the star quickly approaches an equilibrium size that is different 
(and in this case larger) than the one-dimensional model. However, the magnitude 
of this change becomes smaller with resolution. The star is only approximately 
in equilibrium when the coarse grid of $256^3$ zones has a level of 
refinement that jumps by a factor of four. However, even then there is a slight 
uptick in the size toward the event, implying that the numerical stability 
is not guaranteed for arbitrarily long timescales. This result suggests that when 
doing the equilibrium initial models that will former the basis of later calculations,
we should carefully monitor the evolution of the stars before applying any 
artificial damping to cause the merger, to ensure that the merger is due to 
this applied force and not the intrinsic numerical instability of the stars.

@@

A simple dynamical test to verify the gravitation physics in \castro\ is
the case of gravitational free fall. We place two stars on the grid 
in the manner of \autoref{sec:initial_models}. The distance $a$ between 
them corresponds to a chosen orbital period $T$, consistent with the total
system mass $M$, but we disable the rotational source terms so that 
the stars start at rest in an inertial reference frame. 
Thus the stars will simply begin moving toward each other.
As long as the stars remain approximately spherical, the stars can be 
treated as point masses (this only seriously breaks down after the stars
have come into contact). In dimensionless units where $r \to r / a$ and 
$t \to 2\sqrt{2}\pi t / T$, the simple free fall equation of motion governing the
distance $r$ between their centers of mass takes the form:
\begin{equation}
  \ddot{r}(t) = - \frac{1}{2r^2}.
\end{equation}
It is possible to derive a closed-form solution for the evolution time
as a function of separation by starting with the integral formulation,
\begin{equation}
  t(r) = \int_{1}^{r} \frac{dr}{v(r)}.
\end{equation}
The velocity $v$ (in dimensionless units) can be found by noting that 
$\ddot{r} = v\, dv / dr$ and then separating and integrating the equation 
of motion. This yields 
\begin{equation}
  v(r) = \sqrt{\left(\frac{1}{r} - 1\right)}.
\end{equation}
For our problem $0 \leq r \leq 1$, so this is always valid. Integrating, we find
\begin{equation}
  t(r) = \text{arccos}\left(\sqrt{r}\right) + \sqrt{r \left(1 - r\right)}. \label{analyticalFreeFall}
\end{equation}
so that the point of contact would occur at $t = 1$. We actually stop the simulation
at $t = 0.9$, which is when the effects from the extended sizes of the stars
starts to become important. The results of our simulation for our default $256^3$ zone 
uniform grid are shown in \autoref{fig:freefall}. They show excellent agreement
between the theory and the simulation results.

>>>>

The latest developments have resulted in some possible areas of convergence.
\cite{pakmor:2012} performed a merger scenario
with a $1.1\ \msolar$ and $0.9\ \msolar$ setup, with better treatment
of the initial conditions, and indeed found that the merger process
happened over more than ten orbits. Nevertheless, they still determined
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} and \cite{kashyap:2015} were also 
able to find a detonation in similarly massive systems. Notably,
the detonation occurred self-consistently and did not need to be inserted 
by hand. \cite{dan:2012} and \cite{dan:2014} performed a large sweep 
of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many
would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. For now there is preliminary support for the hypothesis
that systems with helium shells (or helium WDs), and very massive carbon-oxygen binaries,
could robustly lead to events resembling SNe Ia.

@@

This is the first in a series of papers designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:implementation} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to demonstrate that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

The Euler equations for hydrodynamics (in the absence of source terms) in conservative form are: 
\begin{align}
  \frac{\partial \rho}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}) \label{eq:euler_density}\\
  \frac{\partial \rho \mathbf{u}}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}\mathbf{u}) - \bm{\nabla}p \label{eq:euler_momentum}\\
  \frac{\partial \rho E}{\partial t} &= -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u}). \label{eq:euler_energy}
\end{align}
Here $\rho$ is the mass density, $\mathbf{u} = (u, v, w)$ is the fluid velocity
vector, $p$ is the pressure, and $E = \mathbf{u}^2 / 2 + e$ is the
total specific energy, where $e$ is the internal (thermal) specific
energy (energy per unit mass).

@@

A related issue that required a code improvement is that in cases of
large density gradients such as the edge of a star, it is possible to
generate negative densities in zones even with the more strongly
limited PPM. This can occur if a region of large density is moving
away from an ambient zone at relatively large speeds; then the net
density flux in the ambient zones can be large enough to unphysically
drag the density below zero. In practice, this will occur at the
trailing edge of a star that is moving across a grid. In such a
situation, there are two main approaches one could take: either
explicitly introduce a positivity-guaranteeing diffusive flux, or
reset the properties of the affected zone. We choose the latter
approach. Even though it is non-conservative, it preserves a
characteristic we value, which is to keep the edge of the stars
relatively sharp, as they physically should be. Since the mass of the
affected zones is typically already fairly low, this should not
seriously affect the energy conservation properties of our
simulation. Our strategy for a reset is as follows: when the density of 
a zone is below a pre-determined density floor (which is typically 
$10^{-5}\ \text{g cm}^{-3}$ for our stellar simulations), we look
at all adjacent zones and find the zone with the highest density.
If it is above the density floor, then we set the field values 
(density, momentum, energy, and temperature) of the
reset zone to be exactly equal to the field values of this 
adjacent zone. If no adjacent zone reaches the density floor, then
the zone is set to the density floor, and given a temperature equal 
to the temperature floor for our simulations (which is typically 
$10^{5}\ \text{K}$ for our stellar simulations). We then recompute 
the thermodynamics to be consistent with these values. The 
velocity of the zone is set to zero. This latter approach will only
occur in very rare situations, and is mainly there as a safeguard.

\castro's approach to adaptive mesh refinement, based on its underlying
\boxlib\ framework, is to refine zones based on certain user-specified
criteria that tag regions of interest for higher spatial
resolution. Data is represented on one of a number of AMR levels,
where each level corresponds to a set of zones at the same resolution,
which covers a subset of the domain covered by the level immediately
below it. We typically call the level 0 grid the \textit{coarse} grid,
which has the lowest spatial resolution. Each finer, higher-level grid
has a higher resolution than the grid below it by some integer factor
$N$, which is restricted to be $N = 2\ \text{or}\ 4$ in \castro. The
zones are strictly contained within the rectangular extent of the
underlying coarser zones (the code is not restricted to representing
only Cartesian geometries, but we use a Cartesian mesh with uniform
spacing in each dimension for the present study). For the time
evolution of the AMR system we use subcycling, where each AMR level is
advanced at a different timestep and a correction step is applied at
the end to synchronize the various levels. The number of
subcycled timesteps is equal to the jump in refinement between levels,
so for example on a grid with three levels and two jumps of four in
refinement, the level 2 zones will have 16 times higher spatial
resolution than the coarse grid and there will be 16 level 2 timesteps
per level 0 timestep.

@@

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
However, we believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

@@

In a typical simulation we place the boundaries of the domain far
enough away from the region containing most of the mass that some
method of approximation to this direct summation is justified. Many
approaches exist in the literature. The original release of \castro\
featured the crudest possible approximation: a monopole prescription,
where the boundary values were computed by summing up all the mass on
the domain and treating it as a point source at the domain
center. This is exactly correct only for a spherically symmetric mass
distribution, and therefore is best suited for problems like
single-star Type Ia supernova simulations (e.g. \cite{malone:2014})
that employ self-gravity. However, for a problem like that of a binary star system
with significant departures from spherical symmetry, this assumption
fails to produce accurate boundary values. This results in a
significant drift of the center of the mass of the system over time,
which will be discussed in \autoref{sec:kepler}.

@@

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, loaded onto the grid by the method described 
in \autoref{sec:implementation}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by \autoref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we will call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in \autoref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. In \autoref{sec:kepler} we 
see that there are no gains to be had by increasing the accuracy further. 
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

@@

We can solve one of these two sources of error by changing the density on  
the grid so that the total mass $M$ is actually what we intend it to be.
The resolution study for this case (the ``normalized sphere'') is also
plotted in \autoref{fig:gravity_convergence}. We still obtain
convergence slightly better than second-order, indicating that we 
have not eliminated the geometrical problem of the mass distribution changing.

@@

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers. We use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations. However, a slight 
difference is that the Coriolis force for each velocity component is coupled to other velocity 
components. If the rotation is about the $z$-axis, then the discrete update to 
$u^{n+1}$ depends on the value of $v^{n+1}$, and vice versa. If we fix the value of 
the time-level $n+1$ quantities after coming out of the hydrodynamics update, there 
would be a slight inconsistency between the $x$ and $y$ components of the velocity.

@@

In all simulations performed in a rotating reference frame, we transform all relevant
quantities back to the inertial reference frame when reporting them in analysis routines 
and visualization (though the data is saved to plotfiles while still in the rotating frame). In particular,
for every zone we adjust the position, momentum, and energy to account for rotation.
If the position is $\mathbf{x}$ in the inertial frame and $\mathbf{x}^\prime$ in 
the rotating frame, and the rotation vector is $\bm{\omega}$, the transformation rules are:
\begin{align}  
  \mathbf{x}(t) &= \mathbf{R}\mathbf{x}^\prime(t) \\
  \mathbf{v}(t) &= \mathbf{v}^\prime(t) + \bm{\omega} \times \left(\mathbf{R} \mathbf{x}^\prime(t)\right)
\end{align}
The rotation matrix $\mathbf{R}$ is:
\begin{equation}
  \mathbf{R} = \mathbf{R}_z({\bm{\theta}}_3) \mathbf{R}_y({\bm{\theta}}_2) \mathbf{R}_x({\bm{\theta}}_1) 
             %% = \left( \begin{array}{ccc} 
             %%     \text{cos}(\theta_2)\, \text{cos}(\theta_3) & 
             %%     -\text{cos}(\theta_2)\, \text{sin}(\theta_3) & 
             %%     \text{sin}(\theta_2) \\
             %%     \text{cos}(\theta_1)\, \text{sin}(\theta_3) + \text{sin}(\theta_1)\, \text{sin}(\theta_2)\, \text{cos}(\theta_3) &
             %%     \text{cos}\theta_1)\, \text{cos}(\theta_3) - \text{sin}(\theta_1)\, \text{sin}(\theta_2)\, \text{sin}(\theta_3) &
             %%     -\text{sin}(\theta_1)\, \text{cos}(\theta_2) \\
             %%     \text{sin}(\theta_1)\, \text{sin}(\theta_3) - \text{cos}(\theta_1)\, \text{sin}(\theta_2)\, \text{cos}(\theta_3) &
             %%     \text{sin}(\theta_1)\, \text{cos}(\theta_3) + \text{cos}(\theta_1)\, \text{sin}(\theta_2)\, \text{sin}(\theta_3) &
             %%     \text{cos}(\theta_1)\, \text{cos}(\theta_2) 
             %%   \end{array} \right)
\end{equation}
where $\mathbf{R}_x$, $\mathbf{R}_y$, and $\mathbf{R}_z$ are the standard rotation matrices about 
the $x$, $y$, and $z$ axes, and $\bm{\theta} = \bm{\omega} t$.

%==========================================================================
% Problem Description and Software Implementation
%==========================================================================

\section{Problem Description and Software Implementation}
\label{sec:implementation}

In this section we describe our white dwarf merger software, and focus in 
particular on the initial white dwarf models (\autoref{sec:initial_models}), 
the initial problem setup (\autoref{sec:initial_state}), and analysis 
(\autoref{sec:analysis}) components.

\subsection{White Dwarf Models}
\label{sec:initial_models}

@@

We map the 1D model onto the 3D Cartesian grid by taking density,
temperature, and composition as the independent variables,
interpolating these to the cell centers, and then calling the equation
of state to initialize the remaining terms. It is possible to interpolate
instead by using pressure instead of temperature, as pressure is more 
closely related to hydrostatic balance, but the EOS we use is so 
insensitive to temperature that this mapping can result in large 
deviations from the isothermal assumption we started with.  The 
interpolation process divides each zone into $n_{\text{sub}}$ 
sub-zones of equal volume for
the purpose of sampling the 1D model, and the sub-zones are added
together to obtain the full zone's state. This
sub-grid-scale interpolation is useful especially near the edge of the star,
where the density falls off rapidly with radius. Typically we take 
$n_{\text{sub}} = 4$.

\subsection{Initial State}
\label{sec:initial_state}

@@

\subsection{Analysis}
\label{sec:analysis}

@@

\subsubsection{Gravitational Waves}
\label{sec:grvatational_waves}

@@

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the origin, 
and evolved it for 200 seconds. Our diagnostic of choice is the effective 
radius of the star, determined by the volume of the grid that has a density 
greater than $10^3\ \text{g cm}^{-3}$ (see \autoref{sec:implementation} 
for details on this measure). This choice of density is intended to 
mark a reasonable outer edge to the star that is not immediately susceptible 
to the numerical errors prevalent near the physical edge of the star. 
\autoref{fig:single_star_static_radius} shows our results at various resolutions. 
As expected, the star quickly approaches an equilibrium size that is different 
(and in this case larger) than the one-dimensional model. However, the magnitude 
of this change becomes smaller with resolution. The star is only approximately 
in equilibrium when the coarse grid of $256^3$ zones has a level of 
refinement that jumps by a factor of four. However, even then there is a slight 
uptick in the size toward the event, implying that the numerical stability 
is not guaranteed for arbitrarily long timescales. This result suggests that when 
doing the equilibrium initial models that will former the basis of later calculations,
we should carefully monitor the evolution of the stars before applying any 
artificial damping to cause the merger, to ensure that the merger is due to 
this applied force and not the intrinsic numerical instability of the stars.

@@

A simple dynamical test to verify the gravitation physics in \castro\ is
the case of gravitational free fall. We place two stars on the grid 
in the manner of \autoref{sec:implementation}. The distance $a$ between 
them corresponds to a chosen orbital period $T$, consistent with the total
system mass $M$, but we disable the rotational source terms so that 
the stars start at rest in an inertial reference frame. 
Thus the stars will simply begin moving toward each other.
As long as the stars remain approximately spherical, the stars can be 
treated as point masses (this only seriously breaks down after the stars
have come into contact). In dimensionless units where $r \to r / a$ and 
$t \to 2\sqrt{2}\pi t / T$, the simple free fall equation of motion governing the
distance $r$ between their centers of mass takes the form:
\begin{equation}
  \ddot{r}(t) = - \frac{1}{2r^2}.
\end{equation}
It is possible to derive a closed-form solution for the evolution time
as a function of separation by starting with the integral formulation,
\begin{equation}
  t(r) = \int_{1}^{r} \frac{dr}{v(r)}.
\end{equation}
The velocity $v$ (in dimensionless units) can be found by noting that 
$\ddot{r} = v\, dv / dr$ and then separating and integrating the equation 
of motion. This yields 
\begin{equation}
  v(r) = \sqrt{\left(\frac{1}{r} - 1\right)}.
\end{equation}
For our problem $0 \leq r \leq 1$, so this is always valid. Integrating, we find
\begin{equation}
  t(r) = \text{arccos}\left(\sqrt{r}\right) + \sqrt{r \left(1 - r\right)}. \label{analyticalFreeFall}
\end{equation}
so that the point of contact would occur at $t = 1$. We actually stop the simulation
at $t = 0.9$, which is when the effects from the extended sizes of the stars
starts to become important. The results of our simulation for our default $256^3$ zone 
uniform grid are shown in \autoref{fig:freefall}. They show excellent agreement
between the theory and the simulation results.
